Ornaments as indicators of social changes in northeastern Taiwan before and after the European colonial period

### 2020-07-01 analysis/paper/paper.Rmd
corrected references and added one para for chronology

One of the most commonly traded types of object in this region were ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. The consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, we explore archaeological ornaments from Kiwulan spanning the pre-European contact period, the period of Spanish and Dutch presence, and the following period of Chinese presence. We address the question of whether indirect colonial influences on the Indigenous populations can be detected through the ornament assemblages.

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (1350-1850 AD, 600-100 BP) and a lower component (650-1150 AD, 1200-800 BP) separated by a sterile deposit spanning c. 150 years [@Chen2007]. The sterility may be related to dry weather, according to pollen analysis, leading to site abandonment [@Lin2015]. These component divisions are based on the differences in color and texture of the deposit and the content of artifacts. Whether these two components represent a culture continuity from the same people is still under debate [@Chiu2004; @Chen2007]. There are 32 radiocarbon ages spanning the two components, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). We focus on the upper component because only this component spans the periods of pre-European contact, European presence, and Chinese presence. In the upper component, all excavation squares in our sampling area show signs of continuous human occupation during each of the three phases. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the radiocarbon dates, excavation depth, consistency of contexts, and types of chronologically diagnostic ceramics such as blue and white porcelains [@Hsieh2009; @Wang2011]. Based on these six units and copies of original excavation records, we have made some refinements of the original chronology to assign contexts into the pre-European, European, and the Chinese periods. The methods we used to devise the new chronology are described in the Supplementary Online Materials.

@@

The Kiwulan ornament data may be interpreted as indicating that Yilan social organization moved from a corporate mode, before the European arrival, to a network mode during European presence. The changes from a less concentrated to a more concentrated distribution of ornaments before and after the presence of Europeans appears consistent with the shift from shared power and wealth to accumulated wealth and monopolization. One possible explanation for the shift could be the long-distance trade network introduced by Europeans. The rarity and the image of colonial power of foreign trade goods resulted in the emergence of competition among ambitious individuals for prestige, wealth, or power through collecting them [@Boone1992; @Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. Resistance could be presented in many forms, but we have no specific evidence from Kiwulan to prefer resistance as the primary mechanism behind the distribution of ornaments over colonial influence. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences. As noted above, this pattern is not a good fit for the ornament distribution at Kiwulan, and adds little value in explaining the shifts in spatial patterns.

@@

Examination of the archaeological record at the peripheries of colonial activity shows how remote Indigenous groups were affected by major European colonial processes [@Trabert2017], and similar impacts can be also seen in modern societies today when adopting western products [@AlGhanim2017; @Jane2015]. Kiwulan in northeastern Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the reach of direct European colonial occupation. The frequency and spatial distribution of personal ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects an increasing use in ornaments in a colonial context. Before European contact, ornaments were traded into local Indigenous societies via the regional exchange network with Chinese merchants, and viewed as prestige goods in the local Indigenous culture. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signaling wealth and privileged trading connections among the inhabitants of Kiwulan. This may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality. This might also indicate an act of intentional resistance to the intrusion of the Europeans by using more ornaments that are symbolic of cultural tradition, but additional evidence is required to confirm this.

>>>>

One of the most commonly traded types of object in this region were ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. The consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, we explore archaeological ornaments from Kiwulan spanning the pre-European contact period, period after the Spanish and Dutch presence, and the following period after the Chinese presence. We address the question of whether indirect colonial influences on the Indigenous populations can be detected through the ornament assemblages.

@@

## Chronology

The chronology of Kiwulan can be divided into two phases represented by a upper component (1350-1850 AD, 600-100 BP) and a lower component (650-1150 AD, 1200-800 BP) based on 32 radiocarbon ages, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). There is a sterile deposit between two components spanning c. 150 years that may be related to dry weather, according to pollen analysis, leading to site abandonment [@Chen2007; @Lin2015]. These component divisions present the differences in color and texture of the deposit, and the content of artifacts such as pottery types. Whether these two components suggest a culture continuity from the same people is still under debate [@Chiu2004; @Chen2007]. We focus on the upper component because this component spans the periods when local residents interacted with the Europeans and the Chinese. In our study area, all excavation squares demonstrate signs of continuous human occupation throughout the upper component. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the types of chronologically diagnostic ceramics, excavation depth, consistency of contexts, and radiocarbon dates [@Hsieh2009; @Wang2011]. To explore the changes in ornament assemblages over time associated with foreign impacts, we assigned the original excavation levels into three phases: before the European contact, after the European presence, and the presence of large Chinese immigrants.

Despite the relatively short period of the European colonization in Taiwan (38-years-long Dutch occupation and 16-years-long Spanish occupation), the indirect impacts of the Europeans through trade network could go beyond the temporal boundaries and last longer. This is because the use of foreign goods in many Indigenous societies often go through the processes of negotiations, resistance, or transformation that a static time frame may not capture the processes well [@Scheiber2012]. Also, we acknowledged that a fine-grained identification of decades of the European presences from the archaeological record is challenging in our research based on some reasons. The current radiocarbon ages that are unable to provide a high resolution of time scale, the circulation of foreign goods might start earlier before the European presences, and the use of the foreign goods in Indigenous societies could last longer after the official end of European colonization [@Mitchell2010]. It is more suitable to use pericolonial approaches that consider a long-term indirect consequences of foreign impacts rather than a focus on specific time frame [@Trabert2018; @Scheiber2012].

Based on these six units and copies of original excavation records, we have made some refinements of the original chronology to assign contexts into the pre-European, European, and the Chinese periods. The methods we used to devise the new chronology are described in the Supplementary Online Materials.

@@

The Kiwulan ornament data may be interpreted as indicating that Yilan social organization moved from a corporate mode, before the European arrival, to a network mode during European presence. The changes from a less concentrated to a more concentrated distribution of ornaments before and after the presence of Europeans appears consistent with the shift from shared power and wealth to accumulated wealth and monopolization. One possible explanation for the shift could be the long-distance trade network introduced by Europeans. The rarity and the image of colonial power of foreign trade goods resulted in the emergence of competition among ambitious individuals for prestige, wealth, or power through collecting them [@Boone1992; @Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012]. 
That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. Resistance could be presented in many forms, but we have no specific evidence from Kiwulan to prefer resistance as the primary mechanism behind the distribution of ornaments over colonial influence. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences. As noted above, this pattern is not a good fit for the ornament distribution at Kiwulan, and adds little value in explaining the shifts in spatial patterns.

@@

Micro-scale analysis of the archaeological record at the peripheries of colonial activity offers a opportunity to understand pervasive but not necessarily dominant European colonial processes at remote Indigenous groups [@Trabert2018]. Similar impacts can be also seen in modern societies today when adopting western products [@AlGhanim2017; @Zheng2015]. Kiwulan in northeastern Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the reach of direct European colonial occupation. The frequency and spatial distribution of personal ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects an increasing use in ornaments in a colonial context. Before European contact, ornaments were traded into local Indigenous societies via the regional exchange network with Chinese merchants, and viewed as prestige goods in the local Indigenous culture. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signaling wealth and privileged trading connections among the inhabitants of Kiwulan. This may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality. This might also indicate an act of intentional resistance to the intrusion of the Europeans by using more ornaments that are symbolic of cultural tradition, but additional evidence is required to confirm this.
